---
title: "Post-matching analysis in *R* using the `mlm` function"
author: "Ben B. Hansen, Mark Fredrickson, Josh Buckner, Josh Errickson, and Peter Solenberger, with embedded Fortran code due to Dimitri P. Bertsekas and Paul Tseng"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Post-matching analysis with mlm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, prompt=TRUE)
```

# The *R* Environment

All the software used in this worksheet is freely available. The *R*
statistical package is installed for you in the lab, but you may
download and install *R* for Windows, Mac, and Linux systems from:
<http://www.r-project.org>.

The following document walks through a common propensity score
matching work-flow in *R*. Example *R* code will appear with a `>`
indicating the command prompt. You may type this code yourself ---
each line is a command to *R*. Output will follow prefaced by
`##`. (In *R*, `#` represents a comment; any command preceded by any
number of `#`'s will not be executed.) For example:

```{r}
2 + 2
```

(Note that when entering the code yourself, do not include the `>` in
your command. Also, for longer lines of code in this document, the
text may wrap onto a second line, with the second line preceded by a
`+` sign. When entering the code yourself, you do not have to wrap the
lines, and do not include the `+`.)

*R* stores data in named variables using the arrow operator:

```{r}
my.variable <- 2 + 2
my.variable * 3
```

# Setup

## Outfitting your *R* with the proper add-ons

*R* add-on packages are available to install directly from *R*:

```{r,eval=FALSE}
install.packages("optmatch")
install.packages("RItools")
```

These commands will ask you to select a CRAN server. Any server will
do.  You may also be asked whether you'd like to set up a "personal
library to install packages into"; if so, answer yes.  (The default
personal library location that *R* will suggest should be OK.) You'll
only need to run these commands the first time you want to use
`optmatch` or `RItools` on a particular computer, or when you install
a new version of *R*.

You'll also need to enable the `mlm` function. It is in an
optmatchExperimental branch, which as the name implies, exists for
testing out new code that will eventually migrate into the main
`optmatch` branch. The following two commands will load in a working copy of `mlm`:

```{r,eval=FALSE}
install.packages("devtools")
devtools::install_github(repo = "markmfredrickson/optmatchExperimental")
```

## Setting up the *R* environment for matching

Attach extension packages that we'll be using for matching and
associated diagnostics:

```{r,echo=FALSE,message=FALSE}
library(optmatch)
library(RItools)
devtools::load_all()
```

```{r,eval=FALSE}
library(optmatch)
library(RItools)
library(optmatchExperimental)
```

You'll do this each time you start a new *R* session and want to run
matching commands.

To load the nuclear plants data, enter

```{r, echo=FALSE}
data(nuclearplants)
```

```{r, eval=FALSE}
data(nuclearplants)
```

To see the first six rows:

```{r}
head(nuclearplants)
```

For more on the variables here, enter

```{r, eval=FALSE}
help("nuclearplants")
```

You can directly access a variable within this data frame as follows.
(Try typing in the commands to see what they do.)

```{r,eval=FALSE}
nuclearplants$pt
table(nuclearplants$pt)
with(nuclearplants, table(pt))
```

The variable you will have just viewed and tabulated, `pt`, is a dummy
for whether the plant was built with "partial turnkey guarantees."
These plants were not comparable to the others in terms of
construction costs.  Let's exclude them for the time being, for
simplicity.  To do this we'll create a data table (in *R* jargon, a
"data frame") of just those observations for which `pt` is 0:

```{r}
nuke.nopt <- subset(nuclearplants, pt == 0)
```

To inspect its first six or last six entries, do

```{r,eval=FALSE}
head(nuke.nopt)
tail(nuke.nopt)
```

To view this as presenting a matching problem, we'll think of plants
built on the site of a previously existing plant (`pr == 1`) as the
treatment group and plants on new sites (`pr == 0`) as comparisons.


# Optimal pair matching and 1:k matching

## Pair matching

To check the number of treated and control plants:

```{r}
table(nuke.nopt$pr)
```

To get the pair match minimizing the mean paired distance on `cap`,
among all collections of 7 non-overlapping pairs, do

```{r}
pairmatch(pr ~ cap, data = nuke.nopt)
```

For a more readable report of who gets matched to whom, type

```{r}
print(pairmatch(pr ~ cap, data = nuke.nopt), grouped = TRUE)
```

For matching on both `date` and `cap`, you'd type `pairmatch(pr ~ cap
+ date, ...)` instead of `pairmatch(pr ~ cap, ...)`. We'll talk later
about how this combines discrepancies on the two variables.  For now,
note the form of the output this command generates: a variable of the
same length as the variables making up `nuke.nopt`, assigning a
distinct name to each matched set.  To fix your intuition, you might
try connecting up below the units that `pairmatch()` has placed in the
same matched sets.

```{r, results="asis", echo=FALSE, warning=FALSE}
library(pander)
a <- with(nuke.nopt, data.frame(
                         Plant=row.names(nuke.nopt),
                         Date=round(date-65, 1),
                         Capacity=round(x=(cap-400),digits=-1))[as.logical(pr),])

b <- with(nuke.nopt, data.frame(
                         Plant=row.names(nuke.nopt),
                         Date=round(date-65, 1),
                         Capacity=round(x=(cap-400),digits=-1))[!as.logical(pr),])

c <- cbind(data.frame(rbind(as.matrix(a), matrix(nrow=nrow(b)-nrow(a), ncol=3))), b)
rownames(c) <- NULL
pandoc.table(c, style="multiline", missing="",
             caption='New-site (left columns) versus existing-site (right columns) plants. "date" is `date-65`; "capacity" is `cap-400`.')
```

For basic summary information about this match, try

```{r, eval=FALSE}
summary(pairmatch(pr ~ cap, data = nuke.nopt))
```

If you've already typed in the `pairmatch(...)` part, you can use the
up-arrow, Home and End keys to avoid having to re-type. Alternatively,
to assign the name "`pm`" to the matching result, do

```{r}
pm <- pairmatch(pr ~ cap, data = nuke.nopt)
```

Now, you can just type `print(pm, grouped = TRUE)` or `summary(pm)`.

The following would give a basic matched analysis of the effect of new
or existing site on construction costs is given with the help of *R*'s
linear modeling function.  In effect, the existing site effect is
estimated as one "way" in a two-way \textsc{anova}, the other "way"
being the factor variable that represents the matching result,
i.e. `pm`.

```{r, eval=FALSE}
summary(lm(cost ~ pr + pm, data = nuke.nopt))
```

## Matching with multiple controls

There are other types of matches you might want to try.  Here's how to
create matched triples (each treatment group unit is matched to two
control group units):

```{r}
tm <- pairmatch(pr ~ cap, controls = 2, data = nuke.nopt)
```

There will be further variations suggested on the slides.

## Did matching work?

It's possible to give the software an impossible list of requirements
for a match.  For instance, try running the following:

```{r, eval=FALSE}
pairmatch(pr ~ cap, controls = 3, data=nuke.nopt)
```

The problem here is that the data don't have 3 comparison units to go
with each treatment unit, since we have 7 treatment units but only 19
comparison units.

Matching can also fail because the distance matrix embodies matching
constraints that are impossible to meet.  In these cases the matching
function will generally run without complaint, although it won't
create any matches.  Here is an example, where the caliper is so
narrow as to forbid all possible matches:

```{r, error=TRUE}
pairmatch(pr ~ cap + cost, caliper=.001, data = nuke.nopt)
```

Behind the scenes, the `caliper` argument restricts how the maximum
distance between matched objects. For example, consider Table 1
above. Plants A and H are 1.3 units apart in `date`. If we assigned
`caliper=1`, they could never be matched because they exceed the
caliper limit.

If before matching you want to remove just the subjects lacking a
counterpart within caliper distance, you can do `pairmatch(...,
remove.unmatchables = TRUE)`.  That won't help with the minuscule
caliper above, but with less extreme calipers it helps you salvage a
few matches.


## How closely did I match?

Getting back to a matching that succeeded, note that `summary()`
reports information about how close the matches are.

```{r}
summary(pm)
```

## Did matching balance the covariate?

Comparing overt biases before and after matching.  An assessment of
the unmatched difference between the groups on `cap` can be had via:

```{r, eval=FALSE}
cap.noadj <- lm(cap ~ pr, data = nuke.nopt)
summary(cap.noadj)
```

The output is suppressed, as most it is not relevant to balance.
This variation hones in on the part that is:

```{r}
summary(lm(cap ~ pr, data = nuke.nopt))$coeff["pr",]
```

(Note again the use of square brackets, `[` and `]`, for specifying
subsets of a matrix.  With *R* one has to carefully distinguish square
brackets, curly brackets and parentheses.)

Here is a parallel calculation that takes the match `pm` into account.

```{r}
summary(lm(cap ~ pr + pm, data = nuke.nopt))$coeff["pr",]
```

The RItools package's `xBalance` function zeroes in on balance, and
facilitiates checking balance on multiple variables at the same time. Here are some examples:

```{r}

xBalance(pr ~ cap + t2, report="all", data=nuke.nopt)
xBalance(pr ~ cap + t2 + strata(pm),##, strata=data.frame("none"="-", "pm"=pm),
         data=nuke.nopt,
         report=c("adj.mean.diffs", "std", "z"))
```

__Exercises.__

1. Compare `pm`, `tm` and the unmatched samples in terms of balance on
`t2`.
1. Compare `pm`, `tm` and the unmatched samples in terms of balance on
`date`.
1. Compare `pm` to Mahalanobis pair matching on `t1` in terms of
balance on `date`.
1. Compare Mahalanobis pair matching on `cap` and `date` to
  Mahalanobis pair matching on `cap`, `date` and each of `t1,t2`.  Add
  the last two variables in one at a time, so that you're comparing a
  total of three matches.  Compare on balance in `cap` and `t2`.

# Using `mlm` for post-matching analysis

Let's attempt to fit a regression model that accounts for the matched
structure. We saw earlier that we can do a basic matched analysis by
adding a fixed effect per matched set.

```{r,eval=FALSE}
summary(lm(cost ~ t1 + pm, data = nuke.nopt))
```

However, as the number of matched sets grows, this can become
unwieldy. The `mlm` function addresses this concern by instead
accounting for the matched structures internally. The syntax mirrors
the syntax for `lm`, and has the additional requirement that one
object on the right-hand side of the formula is an `optmatch` object
(in this case, `pm`).

```{r}
mlm1 <- mlm(cost ~ t1 + pm, data = nuke.nopt)
summary(mlm1)
```

You'll notice that there is no coefficient estimated for `pm`.

The standard errors associated with the coefficients in `mlm1` do not
account for the matched structure either. We can use sandwich
estimators to properly compute them. Sandwich-style estimates of
standard errors are more robust to model misspecification, but are
less efficient than traditional standard error estimates.

(The `sandwich` package should be installed on R by default, if it
appears to not be, you may need to install it using
`install.packages(sandwich)`.)

```{r}
library(sandwich)
sandwich(mlm1)
```

Keep in mind that `sandwich` returns variance estimates, so we need to
take the squareroot of the diagonal components to get the standard
error estimates.

```{r}
sqrt(diag(sandwich(mlm1)))
```